---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, unzip the source file in your Working Directory.

```{r unzip, cache=TRUE}
unzip("activity.zip")
```
Read the source csv file (activity) into  data frame and show some rows.
```{r reading, cache=TRUE}
activity <- read.csv("activity.csv", stringsAsFactors=FALSE)
str(activity)
```

Convert variable date to a Date format using lubridate package. 
Install this package if you don't have it.

```{r date_conversion, message=FALSE}
if(!require(lubridate)){
install.packages("lubridate")
library(lubridate)
}
activity$date <- ymd(activity$date)

str(activity)

```


## What is the average daily activity pattern?

#### 1. Calculate the total number of steps taken per day (ignore the missing values)
```{r daily_total,message=FALSE}
if(!require(dplyr)){
install.packages("dplyr")
library(dplyr)
}
total_day <- activity %>% group_by(date) %>%summarise(total_steps=sum(steps,na.rm=TRUE),na=mean(is.na(steps))) %>% print
```
Visualize the total number of steps per day in a barplot (Package "dplyr" is needed.)

```{r barplot,fig.width=12}
# Barplot
barplot(height = total_day$total_steps,names.arg=total_day$date,cex.names=0.68,las=3,col="red")

# Median - Line & Text
abline(h=median(total_day$total_steps), lty=2,lwd=3, col="black")
text(x = 0,y=median(total_day$total_steps),pos=3,labels = "median")

# Mean - Line & Text
abline(h=mean(total_day$total_steps), lty=2,lwd=3, col="blue")
text(x = 0,y=mean(total_day$total_steps),pos=1,labels = "mean",col="black")
```
(PA1_template_files/figure-html/barplot-1.png)

####  2. Make a histogram of the total number of steps taken each day

Days with missing observations does not appear in a Histogram.

```{r histogram}
# Histogram
total_day <- filter(total_day, na < 1)
hist(total_day$total_steps,col="orange",breaks=20,main="Total steps per day",xlab="Steps per day")

# Median - Line & Text
abline(v=median(total_day$total_steps),lty=3, lwd=2, col="blue")
legend(legend="median","topright",lty=3,lwd=2,bty = "n")
```
(PA1_template_files/figure-html/histogram-1.png)

#### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r summary}
# Mean
mean_steps <- mean(total_day$total_steps,na.rm=TRUE)
# Median
median_steps <- median(total_day$total_steps,na.rm=TRUE)
```
Mean and median of the total number of steps taken per day are `r round(mean_steps,2)` steps and `r median_steps` steps, respectively.

## What is the average daily activity pattern?

#### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r daily, fig.width=12}
# Installation of dplyr was already done before.
library(dplyr,quietly = TRUE)

daily_patterns <- activity %>% group_by(interval) %>% summarise(average=mean(steps,na.rm=TRUE))

# Plot
plot(x = 1:nrow(daily_patterns),y = daily_patterns$average,type = "l",
col = "red", xaxt = "n",xlab="Intervals", 
ylab = "Average across all days")
axis(1,labels=daily_patterns$interval[seq(1,288,12)],
at = seq_along(daily_patterns$interval)[seq(1,288,12)])
```
(PA1_template_files/figure-html/daily-1.png)

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r daily2}
max_numb_steps_interval <- filter(daily_patterns,average==max(average))
```

Interval **"`r max_numb_steps_interval$interval`"** contains on average the maximum number of steps (**`r round(max_numb_steps_interval$average,2)` steps**).

## Imputing missing values

#### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
na_number <- sum(is.na(activity$steps))
na_number
percentage_na <- mean(is.na(activity$steps))
percentage_na
```
Missing values amounts to **`r na_number` ** (what is **`r round(percentage_na*100,1)`** % of total observations).

#### 2. Devise a strategy for filling in all of the missing values in the dataset

Missing values will be filled by the average number of steps in particular 5-minutes interval.

#### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r na_imputing, cache=TRUE}
noNAs <- numeric(nrow(activity))
for (i in 1:nrow(activity))
{
  if (is.na(activity[i,"steps"])==TRUE)
  {
  noNAs[i]<-filter(daily_patterns,interval==activity[i,"interval"]) %>% select(average)
  } 
  else
  {
  noNAs[i]<-activity[i,"steps"]
  }
  
}
activity_noNAs<-mutate(activity,steps_no_NAs=noNAs)
head(activity_noNAs)
```

Impact of NAs   
```{r test_NAs}
check <- filter(activity_noNAs,!is.na(steps)) %>% mutate(ok = (steps==steps_no_NAs))
mean(check$ok)

```

#### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r histogram_no_NAs}
total_day_noNAs <- activity_noNAs %>% mutate(steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(date) %>% summarise(total_steps=sum(steps_no_NAs))

# Histogram
hist(total_day_noNAs$total_steps,col="blue",breaks=20,main="Total Steps per day",xlab="Steps Qty")

# Median
abline(v=median(total_day$total_steps),lty=3, lwd=2, col="black")
legend(legend="Median","topright",lty=3,lwd=2,bty = "n")
```

```{r summary_no_NAs}

summary(total_day_noNAs$total_steps)
```
(PA1_template_files/figure-html/histogram_no_NAs-1.png)

Including NAs should affect the final results, our mean increased its value while median decreased.
Missing values always affect mean & median. (Note: Don't confuse a missing value with a zero value)

## Are there differences in activity patterns between weekdays and weekends?

#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r weekday}

#Installation of dplyr was already done before.
library(lubridate)
is_weekday <-function(date){
  if(wday(date)%in%c(1,7)) result<-"Weekend"
  else
    result<-"Weekday"
  result
}

activity_noNAs <- mutate(activity_noNAs,date=ymd(date)) %>% mutate(day=sapply(date,is_weekday))

table(activity_noNAs$day)
```

#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


```{r weekend_comparison, cache=TRUE, fig.width=10}
# ggplot2 packae is required.
if(!require(ggplot2)){
  install.packages("ggplot2")
  library(ggplot2)
}

daily_patterns <- activity_noNAs %>% mutate(day=factor(day,levels=c("Weekend","Weekday")),steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(interval,day) %>% summarise(average=mean(steps_no_NAs))

# Line chart
qplot(interval,average,data=daily_patterns,geom="line",facets=day~.)
```
(PA1_template_files/figure-html/weekend_comparison-1.png)
